---
title: "ADA PROJECT"
author: "Septer"
date: "2024-11-24"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r}
pacman::p_load(gridExtra, knitr, kableExtra, mosaic, xtable, pscl, multcomp, pander, MASS, tidyverse, lmtest, readr, sandwich, broom)
```

```{r}
# introduce the data
Thai <- read.csv("THA2021.csv")
```

```{r}
#Listing available Variables
colnames(Thai)
```
## Data Management
```{r}
#Select required variables
library(dplyr)

Thai_R12 <- Thai[, c("q1", "q2", "q22", "q24", "q27", "q28", "q34", "q40", "q49", "q54", "q56", "q70")]

#edit variable names
Thai_R <- Thai_R12 %>%
  rename(Age = q1, Gender = q2, Loneliness = q22, Suicide = q24, Number_of_friends = q27, Smoking = q28, Alcohol = q34,  Drug_use = q40, Physical_Act =q49, School_Environ = q54, Parental_support = q56, Bullying = q70)

#remove missing Data
Thai_R_Final <- Thai_R %>%
  mutate(across(c(Age, Gender, Loneliness, Suicide, Number_of_friends, Smoking, Alcohol, Drug_use, Physical_Act, School_Environ, Parental_support, Bullying), 
                ~ na_if(as.character(.), "Missing"))) %>%
 filter(complete.cases(Age, Gender, Loneliness, Suicide, Number_of_friends, Smoking, Alcohol, Drug_use, Physical_Act, School_Environ, Parental_support, Bullying)) -> Thai_R_Final
```
```{r}
colnames(Thai_R_Final)
```

##rename data codes back to original code names
```{r}

 # Set factor levels for Age

Thai_R_Final <- Thai_R_Final %>%
  mutate(
    Age = as.numeric(as.character(Age)),  # Convert Age to numeric
    Age = factor(case_when(
      Age <= 3 ~ "Under_13",
      Age >= 4 & Age <= 5 ~ "14_15",
      Age >= 6 & Age <= 7 ~ "16_17",
      Age >= 8 ~ "Over_18"
    ), levels = c("Under_13", "14_15", "16_17", "Over_18")),

    # Set factor levels for Gender
     Gender = factor(case_when(
      Gender == 1 ~ "Male",
      Gender == 2 ~ "Female"
    ), levels = c("Female", "Male")),  
   
    # Set factor levels for loneliness
    Loneliness = factor(case_when(
      Loneliness == 1 ~ "Never",
      Loneliness == 2 ~ "Rarely",
      Loneliness == 3 ~ "Sometimes",
      Loneliness == 4 ~ "Mosttimes",
      Loneliness == 5 ~ "Always"
    ), levels = c("Never", "Rarely", "Sometimes", "Mosttimes", "Always")),  
    
    # Set factor levels for suicide
     Suicide = factor(case_when(
      Suicide == 1 ~ "Yes",
      Suicide == 2 ~ "No"
    ), levels = c("No", "Yes")),  

    # Set factor levels for number of friends
    Number_of_friends = factor(case_when(
      Number_of_friends == 1 ~ "0",
      Number_of_friends == 2 ~ "1",
      Number_of_friends == 3 ~ "2",
      Number_of_friends == 4 ~ ">3"
    ), levels = c("0", "1", "2", ">3")),  
    
     Alcohol = as.numeric(as.character(Alcohol)),
     Alcohol = factor(case_when(
      Alcohol == 1 ~ "No",
      Alcohol >= 2  ~ "Yes",
    ), levels = c("No", "Yes")),

    # Set factor levels for drug use 
    Drug_use = factor(case_when(
      Drug_use == 1 ~ "No",
      Drug_use >= 2  ~ "Yes",
    ), levels = c("No", "Yes")),  
    
     # Set factor levels for Parental_support
      Parental_support = factor(case_when(
      Parental_support == 1 ~ "Never",
      Parental_support == 2 ~ "Rarely",
      Parental_support == 3 ~ "Sometimes",
      Parental_support == 4 ~ "Mosttimes",
      Parental_support == 5 ~ "Always"
    ), levels = c("Never", "Rarely", "Sometimes", "Mosttimes", "Always")),
   
     # Set factor levels for Bullying
      Bullying = factor(case_when(
      Bullying == 1 ~ "Yes",
      Bullying == 2 ~ "No"
    ), levels = c("No", "Yes")),  
  )

```

## Data cleaning Flow Chart
```{r}
library(DiagrammeR)

# Create the flowchart
DiagrammeR::grViz("
digraph flowchart {
  graph [layout = dot, rankdir = TB]
  
  # Define nodes
  node [shape = oval, style = filled, fillcolor = lightblue]
  start [label = 'Introduce Data\\nThai']
  final [label = 'Final Cleaned Dataset\\nThai_R_Final']
  
  node [shape = rectangle, style = filled, fillcolor = lightgoldenrod]
  read_data [label = 'Initial Data\n132 variables\n5,661 Observations']
  select_vars [label = 'Select Relevant Variables:\\n8 Variables\n5,661 Observations']
  rename_vars [label = 'Rename Variables:\\n8 Variables\n5,661 Observations']
  remove_missing [label = 'Remove Missing Data:\\n8 Variables\n4,547 Observations']
  
  # Connect nodes
  start -> read_data -> select_vars -> rename_vars -> remove_missing -> final
}
")

```

## Descriptive Tables
```{r}
table1::table1(~ Age + Gender +  Loneliness + Number_of_friends + Alcohol + Drug_use + Parental_support  | Suicide, data = Thai_R_Final)
```
# Interpretations
Age
The age distribution reveals a higher concentration of participants in the 12_13 and 14_15 age groups, with 28.9% and 38.2% of the total sample, respectively. The 16_17 and Over_18 groups represent a smaller percentage (21.3% and 11.5%, respectively).

Gender
Females make up a larger proportion of the Yes group (74.3%) compared to males (25.7%). Overall, females represent 57.9% of the total sample. This indicates that females in this dataset may have higher rates of experiencing the outcome compared to males

Loneliness
The data on loneliness shows that those who answered "Yes" report higher levels of loneliness, particularly in the Mosttimes and Always categories, which account for 29.9% and 17.8%, respectively. In contrast, the No group primarily reports lower levels of loneliness, particularly in the Never and Rarely categories (28.2% and 29.0%, respectively).

Alcohol and Substance use
The data reveals higher rates of alcohol use, and drug use in the Yes group, with 58.7%, and 89.4% reporting these behaviors, respectively. 
The No group has lower percentages, with 23.0% 41.8% consuming alcohol, and 92.6% engaging in drug use.

Parental Support
Parental support follows a similar pattern, with a higher percentage of the Yes group (39.9%) reporting a lack of parental support, while the No group has 25.9% in the "Never" category.

## Visualize distribution  Bar Graphs & Charts
```{r}
library(ggplot2)
library(dplyr)
## distribution by Age
# Create a  Age data frame
data <- data.frame(
  Category = c("Under_11", "12_13", "14_15", "16_17", "Over_18"),
  No = c(3, 1072, 1456, 833, 464),
  Yes = c(1, 244, 282, 135, 57)
)

# Reshape data for plotting
data_long <- data %>%
  gather(key = "Suicide_Ideation", value = "Count", -Category)

# Plot
ggplot(data_long, aes(x = Category, y = Count, fill = Suicide_Ideation)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Age Distribution by Suicide_Ideation", x = "Age Group", y = "Count") +
  theme_minimal()
```
# Interpretations
The age distribution reveals a higher concentration of participants in the 12_13 and 14_15 age groups, with 28.9% and 38.2% of the total sample, respectively. The 16_17 and Over_18 groups represent a smaller percentage (21.3% and 11.5%, respectively). The Yes group tends to be slightly younger, with a higher proportion in the 12_13 and 14_15 categories


```{r}
## Distribution by Gender

# Create the data for gender and 'Yes' outcome (from your table example)
gender_data <- data.frame(
  Gender = c("Female", "Male"),
  Yes = c(534, 185),
  Total = c(2634, 1913)
)

# Calculate the percentage of "Yes" for each gender
gender_data$Percentage <- gender_data$Yes / gender_data$Total * 100

# Plot the pie chart
ggplot(gender_data, aes(x = "", y = Percentage, fill = Gender)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), position = position_stack(vjust = 0.5)) +
  labs(title = "Gender Distribution for 'Yes' Outcome") +
  theme_void() +
  scale_fill_manual(values = c("maroon", "darkgreen"))


```
#Interepretations
Females make up a larger proportion of the Yes group (74.3%) compared to males (25.7%). Overall, females represent 57.9% of the total sample. This indicates that females in this dataset may have higher rates of experiencing the outcome compared to males, suggesting a gender-based difference in mental health or behavior outcomes.
```{r}
# Loneliness data example
loneliness_data <- data.frame(
  Loneliness = c("Never", "Rarely", "Sometimes", "Mosttimes", "Always"),
  No = c(1080, 1112, 1160, 319, 157),
  Yes = c(66, 98, 212, 215, 128)
)

# Reshape data for plotting
loneliness_data_long <- loneliness_data %>%
  gather(key = "Outcome", value = "Count", -Loneliness)
# Set the factor levels for the Loneliness variable in the desired order
loneliness_data_long$Loneliness <- factor(loneliness_data_long$Loneliness, 
                                          levels = c("Never", "Rarely", "Sometimes", "Mosttimes", "Always"))

# Plot
ggplot(loneliness_data_long, aes(x = Loneliness, y = Count, fill = Outcome)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Loneliness Distribution by Outcome", x = "Loneliness Level", y = "Count") +
  theme_minimal()

```

```{r}
# Reshape data for plotting, focusing on the 'Yes' outcome
loneliness_data_long <- loneliness_data %>%
  gather(key = "Outcome", value = "Count", -Loneliness) %>%
  filter(Outcome == "Yes")  # Filter to include only 'Yes' outcome

# Set the factor levels for the Loneliness variable in the desired order
loneliness_data_long$Loneliness <- factor(loneliness_data_long$Loneliness, 
                                          levels = c("Never", "Rarely", "Sometimes", "Mosttimes", "Always"))


# Plot_2
ggplot(loneliness_data_long, aes(x = Loneliness, y = Count, fill = Outcome)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Loneliness Distribution with 'Yes' Outcome", x = "Loneliness Level", y = "Count") +
  theme_minimal()



```
#interpretations
The data on loneliness shows that those who answered "Yes" report higher levels of loneliness, particularly in the Mosttimes and Always categories, which account for 29.9% and 17.8%, respectively. In contrast, the No group primarily reports lower levels of loneliness, particularly in the Never and Rarely categories (28.2% and 29.0%, respectively).


##Model 1 Total Effect
```{r}
# Recode 'Yes' and 'No' to 1 and 0
Thai_R_Final $suicide <- ifelse(Thai_R_Final$Suicide == "Yes", 1, 0)


#Perform a general linear model 
model_1 <- glm (Suicide ~ Loneliness + Age + Gender + Number_of_friends +  Parental_support, 
             family = binomial(link = "logit"), data = Thai_R_Final)

summary(model_1)
```

```{r}

#confidence intervals
broom::tidy(model_1, exponentiate = TRUE, conf.int = TRUE)

 #Tidy the model and round specific columns to 2 decimal places
broom::tidy(model_1, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(
    estimate = round(estimate, 2),
    std.error = round(std.error, 2),
    statistic = round(statistic, 2),
    conf.low = round(conf.low, 2),
    conf.high = round(conf.high, 2)
  )
```
##Model 2 Direct effect
```{r}
# Recode 'Yes' and 'No' to 1 and 0
Thai_R_Final $suicide <- ifelse(Thai_R_Final$Suicide == "Yes", 1, 0)


#Perform a general linear model 
model_2 <- glm (Suicide ~ Loneliness + Age + Gender + Number_of_friends +  Parental_support + Alcohol + Drug_use, 
             family = binomial(link = "logit"), data = Thai_R_Final)

summary(model_2)
```
```{r}
broom::tidy(model_2, exponentiate = TRUE, conf.int = TRUE)

 #Tidy the model and round specific columns to 2 decimal places
broom::tidy(model_2, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(
    estimate = round(estimate, 2),
    std.error = round(std.error, 2),
    statistic = round(statistic, 2),
    conf.low = round(conf.low, 2),
    conf.high = round(conf.high, 2)
  )
```



# Model Visualization

```{r}
# Load required library
library(forestplot)

# Create a data frame for the input data
data <- data.frame(
  term = c("(Intercept)", "LonelinessRarely", "LonelinessSometimes", 
           "LonelinessMosttimes", "LonelinessAlways", "Age14_15", 
           "Age16_17", "AgeOver_18", "GenderMale", 
           "Number_of_friends1", "Number_of_friends2", "Number_of_friends>3",
           "Parental_supportRarely", "Parental_supportSometimes", 
           "Parental_supportMosttimes", "Parental_supportAlways"),
  estimate = c(0.17, 1.41, 2.66, 9.52, 10.67, 0.78, 0.68, 0.55, 0.46, 
               1.21, 1.02, 0.87, 0.81, 0.63, 0.45, 0.49),
  lower = c(0.11, 1.01, 2.00, 7.02, 7.53, 0.63, 0.53, 0.39, 0.38, 
            0.79, 0.67, 0.62, 0.63, 0.50, 0.32, 0.35),
  upper = c(0.26, 1.96, 3.59, 13.05, 15.23, 0.95, 0.88, 0.76, 0.56, 
            1.85, 1.55, 1.22, 1.03, 0.79, 0.61, 0.66)
)

# Prepare the labels
table_data <- cbind(data$term)

# Create the forest plot
forestplot(
  labeltext = table_data,
  mean = data$estimate,
  lower = data$lower,
  upper = data$upper,
  zero = 1,  # Add a vertical reference line at 1
  xlab = "Log Odds",
  lineheight = unit(1.5, "lines"),  # Line height for spacing
  title = "Model Forest Plot",
  col = fpColors(box = "royalblue", line = "darkblue", summary = "royalblue")
)

```


#Interpretations
Loneliness:
There is a clear positive association in the impact of loneliness on the outcome of interest, with higher levels of reported loneliness associated with higher OR of the outcome. 
Those who report feeling lonely "Rarely" have 1.41 times (95% CI: 1.01–1.96), “Sometimes” have 2.66 times (CI: 2.00–3.59), “Most times” have 9.52 times (CI: 7.02–13.05), and “Always have” 10.67 times (CI: 7.02–13.05) times the odds of having the outcome compared to those who never feel lonely. 
All the levels were significant with P<0.05 for the 4 tests.


Age 14–15 (OR = 0.78, 95% CI: 0.63–0.95, p = 0.016): Adolescents aged 14–15 have a 22% lower likelihood of suicidal ideation compared to the reference group. This result is significant.

Age 16–17 (OR = 0.68, 95% CI: 0.53–0.88, p = 0.003): Older adolescents (16–17) show a 32% reduced likelihood of suicidal ideation compared to the reference group. The protective effect strengthens as age increases.

Age Over 18 (OR = 0.55, 95% CI: 0.39–0.76, p < 0.001): Young adults over 18 years old have a 45% reduced likelihood of suicidal ideation. 
This suggests that older individuals may develop coping mechanisms or have fewer stressors associated with suicidal thoughts.

Gender
Male (OR = 0.46, 95% CI: 0.38–0.56, p < 0.001): Males are 54% less likely to report suicidal ideation compared to females. This aligns with evidence that females often report higher rates of suicidal ideation despite males being more likely to die by suicide.

Number of Friends:
1 Friend (OR = 1.21, 95% CI: 0.79–1.85, p = 0.378): Having one friend is associated with a non-significant 21% increase in the odds of suicidal ideation compared to the reference group (e.g., no friends).

2 Friends (OR = 1.02, 95% CI: 0.67–1.55, p = 0.932): Having two friends has no significant association with suicidal ideation.

More than 3 Friends (OR = 0.87, 95% CI: 0.62–1.22, p = 0.399): Having more than three friends slightly reduces the likelihood of suicidal ideation, but this result is not statistically significant

Parental support
Rarely Supported (OR = 0.81, 95% CI: 0.63–1.03, p = 0.081): Rare parental support shows a non-significant trend towards reduced suicidal ideation.

Sometimes Supported (OR = 0.63, 95% CI: 0.50–0.79, p < 0.001): Occasional parental support reduces the likelihood of suicidal ideation by 37%, showing a significant protective effect.

Mostly Supported (OR = 0.45, 95% CI: 0.32–0.61, p < 0.001): Those with frequent parental support have a 55% reduced risk of suicidal ideation, indicating a strong protective factor.

Always Supported (OR = 0.49, 95% CI: 0.35–0.66, p < 0.001): Consistent parental support halves the likelihood of suicidal ideation, further highlighting the importance of a supportive environment.

#3Check for Multicolinearity
```{r}
car::vif(model_1)
```
# interpretation
In summary, based on the GVIF results for loneliness, number of friends, and drug use in my model, I can conclude that collinearity does not pose a significant issue. The variance in my coefficients does not appear to be substantially inflated due to the correlations among these predictors.


